---
title: "DMI - Hands on 1: Proteomic and Metabolomic Characterization of COVID-19 Patient Sera"
subtitle: "Reproducible Statistical Analysis and Visualization"
author: ' Maria Pau Pijoan Manresa (mariapau.pijoan01@estudiant.upf.edu)'
date: 'Last update: `r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
#install.packages("knitr") #can execute R code inside the R markdown documents
#knitr::opts_chunk$set(echo = TRUE) #como se muestra el codigo, i TRUE significa que el cdigo R se muestra en el documento final (html, PDF,word)
#if (!require("knitr")) install.packages("knitr", repos = "http://cran.us.r-project.org")
library(knitr)
```

# Abstract

This report presents a reproducible statistical analysis of proteomic and metabolomic data derived from COVID-19 patient sera. The objective of the exercise was to replicate selected results from a published study, perform exploratory data analysis (EDA) and evaluate biomarker patterns associated with disease severity. Using R, the dataset was cleaned, reformatted and statistically summarized to reproduce descriptive tables and graphical outputs reported in the original manuscript. Supplementary figures were reconstructed using grouped comparisons and non-parametric statistical testing. Additionally, the influence of extreme values was assessed through systematic outlier detection and sensitivity analysis. Finally, a heatmap of standardized biomarkers was generated to explore multivariate clustering patterns across clinical groups. The results confirm that severe COVID-19 cases are characterized by elevated inflammatory and metabolic markers and reduced lymphocyte counts. This exercise demonstrates reproducible research practices, critical evaluation of published findings and effective visualization of complex biomedical data.

# Introduction

This exercise focuses on the reproducible analysis of proteomic and metabolomic data from COVID-19 patient sera. The dataset, derived from a published study [**Proteomic and Metabolomic Characterization of COVID-19 Patient Sera**](https://www.sciencedirect.com/science/article/pii/S0092867420306279), includes healthy individuals, non-COVID-19 patients and COVID-19 patients classified by disease severity. The aim was to replicate selected results from the original manuscript and explore biomarker patterns associated with disease progression.
The analysis was conducted in R and included data cleaning, variable recoding and exploratory data analysis to understand distributions and group differences. Published summary tables and supplementary figures were reconstructed using descriptive statistics and non-parametric group comparisons. The impact of extreme values was assessed through outlier detection and a standardized heatmap was generated to explore multivariate biomarker patterns.

Overall, this exercise emphasizes reproducibility, statistical interpretation, and effective visualization of biomedical data.

# Exercise 1 – Reproducing table 1

For this part, it is essential to download and load Supplementary Table 1 (“Additional Demographical and Baseline Characteristics of COVID-19 Patients and Control Groups”) from the supplementary materials of the publication.

## Data cleaning

```{r}
# -----------------------------
# Final code to reproduce Table 1 structure using table1
# -----------------------------

#install.packages(c("readxl","dplyr","stringr","readr","lubridate","table1","janitor","htmltools"))

library(readxl)
library(dplyr)
library(stringr)
library(readr)
library(lubridate)
library(table1)
library(janitor)
library(htmltools)

# Read data
df <- read_excel("../1-s2.0-S0092867420306279-mmc1.xlsx", sheet = 2)
df <- clean_names(df)

# Build the main Table-1 group column for all rows first 
df <- df %>%
  mutate(
    group = case_when(
      group_d == 0 ~ "Healthy Control",
      group_d == 1 ~ "Non-COVID-19",
      group_d == 2 ~ "Non-severe",
      group_d == 3 ~ "Severe",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(group))  # removes any unexpected codes

# Clean variables used in table 1
df <- df %>%
  mutate(
    # sex coding (1=Male, 0=Female)
    sex = case_when(
      sex_g == 1 ~ "Male",
      sex_g == 0 ~ "Female",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Male","Female")),

    # numeric age
    age = parse_number(as.character(age_year)),

    # numeric BMI ( make it numeric, not character)
    bmi = na_if(as.character(bmi_h), "/") %>%
      str_replace_all(",", ".") %>%
      str_replace_all("[^0-9.\\-]", "") %>%
      na_if("") %>%
      parse_number(),

    # dates + time from onset to admission
    onset_date = as.Date(onset_date_f),
    admission_date = as.Date(admission_date),
    onset_to_admission = as.numeric(admission_date - onset_date),

    # progression date is stored as Excel serial numbers like 43849
    prog_num = suppressWarnings(as.numeric(na_if(as.character(date_of_progression_to_severe_state), "/"))),
    prog_date = as.Date(prog_num, origin = "1899-12-30"),

    # time from admission to severe (only meaningful for Severe)
    admission_to_severe = as.numeric(prog_date - admission_date)
  )

# Add COVID-19 Total rows (duplicate Non-severe + Severe and relabel)
df_total <- df %>%
  filter(group %in% c("Non-severe", "Severe")) %>%
  mutate(group = "COVID-19 Total")

# Final dataset with 5 columns for Table 1
df5 <- bind_rows(df, df_total)

df5$group <- factor(
  df5$group,
  levels = c("Healthy Control","Non-COVID-19","COVID-19 Total","Non-severe","Severe")
)

# sanity check of Ns 
print(df5 %>% count(group))

# Formatting functions
my.render.cont <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("")
  m <- mean(x); s <- sd(x)
  med <- median(x)
  q <- quantile(x, c(.25,.75))
  r <- range(x)
  sprintf("Mean ± SD: %.1f ± %.1f\nMedian (IQR): %.1f (%.1f–%.1f)\nRange: %.1f–%.1f",
          m, s, med, q[1], q[2], r[1], r[2])
}
my.render.cat <- function(x) render.categorical.default(x)

```

## Table construction

```{r}
# Produce Table 1 and save as HTML
t1 <- table1(
  ~ sex + age + bmi + onset_to_admission + admission_to_severe | group,
  data = df5,
  render.continuous = my.render.cont,
  render.categorical = my.render.cat,
  render.missing = NULL, #Removes the "Missing" rows
  overall = FALSE
) 

t1

htmltools::save_html(t1, "table1.html")

```

## Comparison with published results

Discrepancies between the reproduced table and the published table 1:

- Small differences are observed in some summary statistics, for instance the mean or median and IQR for age, BMI and time-related variables. These differences are likely due to rounding conventions, differences in handling missing values or slightly different inclusion criteria applied by the authors, like excluding specific observations or use of pre-cleaned datasets.
- Furthermore, larger discrepancies are observed in the severe COVID-19 subgroup, especially for age, BMI and time-related variables, showing lower values than those reported in the published table. One possible explanation is that the supplementary dataset includes all recorded severe cases, whereas the published analysis may have excluded specific observations, like patients with incomplete records or implausible dates. 
- The reproduced table includes a small negative value in the range of time from admission to severe disease. This likely arises from inconsistencies or recording differences in date fields within the supplementary data and may reflect same-day events or data-entry timing issues that were corrected or excluded in the published analysis.
- Also, several variables reported in the original table, such as symptoms chest, CT findings, comorbodities, oxygenation index and treatments, are not present in the supplementary excel file. Consequently, these rows could not be reproduced and were appropriately ommited. 

All in all, the reproduced table matches the published table 1 as closely as possible, given the available date. The remaining discrepancies are minor and can be reasonably explained by differences in data preprocessing, rounding, missing-data handling and incomplete variable availability in the supplementary dataset. 


# Exercise 2 – Supplementary figure 1

## Figure reproduction

```{r}

#Descriptive statistics for key variables
df_new_variables <- df[, c(19:22, 24:31)]
summary(df_new_variables)

#install.packages("ggplot2")
library(ggplot2)
library(dplyr)

# Sex bar plot

sex_bar_plot <- ggplot(df5, aes(x=sex, fill = sex)) + geom_bar() + labs(
    title = "Male vs Female",
    x = "Sex",
    y = "Count"
  )

sex_bar_plot
# Age histogram 
age_histogram <- ggplot(df5, aes(x = age)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = "Age distribution",
    x = "Age (years)",
    y = "Count"
  )

age_histogram

# Bmi histogram 
bmi_histogram <- ggplot(df5, aes(x = bmi)) +
  geom_histogram() +
  labs(
    title = "Body Mass Index distribution",
    x = "BMI",
    y = "Count"
  )

bmi_histogram

# Onset to admission 
onset_to_admission_histogram <- ggplot(df5, aes(x = onset_to_admission)) +
  geom_histogram() +
  labs(
    title = "Time from Onset to Admission",
    x = "Days",
    y = "Count"
  )

onset_to_admission_histogram

# Admission_to_severe 
admission_to_severe_histogram <- ggplot(df5, aes(x = admission_to_severe)) +
  geom_histogram() +
  labs(
    title = "Time from Admission to Severe",
    x = "Days",
    y = "Count"
  )

admission_to_severe_histogram 


#Correlation plot
# install.packages("corrplot")
library(corrplot)
#Calculate correlation matrix for numeric variables
# use = "complete.obs" to handle any missing (NA) values
df_corr_ready <- df_new_variables %>% mutate(crp_i_mg_l = as.numeric(as.character(crp_i_mg_l))) #sapply(df_corr_ready, class) #to see if the chracter columns have changed to numeric (becasue crp was considered characters)

cor_matrix <- cor(df_corr_ready, use = "complete.obs")


corrplot(cor_matrix,
         method = "color",
         type = "upper",
         addCoef.col = "black", #shows the correlation coeff
         tl.col = "black", #text label color
         tl.srt = 45, #rotate labels for readibility
         diag = FALSE, #Hide the diagonal
         number.cex = 0.6,    # Shrinks the correlation numbers
         tl.cex = 0.7,        # Shrinks the text labels (variable names)
         cl.cex = 0.7)        # Shrinks the color legend scale on the right

# GROUP COMPARISONS

#group is the COVID status!

#Sex by group 
sex_by_group <- ggplot(df5, aes(x = group, fill = sex)) +
  geom_bar(position = "fill") +
  labs(
    title = "Sex distribution by group",
    x = "Group",
    y = "Proportion"
  )

sex_by_group 

#Age by group 
age_by_group <- ggplot(df5, aes(x = group, y = age)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Age by group",
    x = "Group",
    y = "Age (years)"
  )


age_by_group

#BMI by group 
bmi_by_group <- ggplot(df5, aes(x = group, y = bmi)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Body mass index by group",
    x = "Group",
    y = "BMI"
  )

bmi_by_group

# Time variables by severity 
#Time from symtpom onset to hospital admission was explored only among COVID-19 patients, as this variable was not defined for healthy controls or non-COVID individuals. 
covid_df <- df5 %>%
  filter(group %in% c("Non-severe", "Severe"))

ggplot(covid_df, aes(x = group, y = onset_to_admission, fill = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_manual(
    values = c(
      "Non-severe" = "#4DAF4A",  # green
      "Severe" = "#E41A1C"       # red
    )
  ) +
  labs(
    title = "Time from onset to admission by COVID-19 severity",
    x = "COVID-19 severity",
    y = "Days"
  ) +
  theme(legend.position = "none")


```

## Outlier detection

Seeing the plot above, we can observe that there are many dots high above the rest of the group. To deal with them, we can: Remove, Transform (log the data pulls extreme outliers closer to the mean, making the distribution more normal), or use robust statistics ( use median and IQR instead of mean and SD. By using the median, this is more robust because it is not pulled away by a single extreme value)

```{r}

# REMOVAL ---------------
#install.packages("ggstatsplot")
#install.packages("tidyverse")
#install.packages("patchwork")
library(tidyverse)
library(ggstatsplot)
library(patchwork)
library(readxl)
library(janitor)
library(ggplot2)
library(ggpubr)

mydata<- read_excel("../1-s2.0-S0092867420306279-mmc1.xlsx", sheet = 2)
mydata <- clean_names(mydata)

# Define the variables in the EXACT order to appear in the grid

ordered_vars <- c(
  "wbc_count_109_l", "lymphocyte_count_109_l", "monocyte_count_109_l", 
  "platelet_count_109_l", "crp_i_mg_l", "alt_j_u_l", 
  "ast_k_u_l",          
  "ggt_l_u_l",          
  "tbil_m_mmol_l",       
  "dbil_n_mmol_l",       
  "creatinine_mmol_l",   
  "glucose_mmol_l"       
)

# Data Processing Pipeline
mydata_ordered <- mydata %>%
  # Grouping logic
  mutate(
    group = case_when(
      group_d %in% c(0, 1) ~ "Non-COVID-19",
      group_d == 2 ~ "Non-severe",
      group_d == 3 ~ "Severe",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Non-COVID-19", "Non-severe", "Severe"))
  ) %>%
  filter(!is.na(group)) %>%
  
  # Convert measurements to numeric
  mutate(across(all_of(ordered_vars), as.numeric)) %>%
  select(group, all_of(ordered_vars)) %>%
  
  # Pivot to long format
  pivot_longer(cols = -group, names_to = "measurement", values_to = "value") %>%
  
  # Convert measurement to a factor using the order defined above
  mutate(measurement = factor(measurement, levels = ordered_vars)) %>%
  
  # Outlier cleaning (per group and per marker)
  group_by(group, measurement) %>%
  mutate(
    q1 = quantile(value, 0.25, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    up = q3 + (1.5 * iqr),
    low = q1 - (1.5 * iqr),
    cleaned_value = ifelse(value > up | value < low, NA, value)
  ) %>% 
  ungroup()
# Define the comparisons want to show (for instance every group against each other)
my_comparisons <- list( 
  c("Non-COVID-19", "Non-severe"), 
  c("Non-severe", "Severe"), 
  c("Non-COVID-19", "Severe") 
)
```

## Cleaned figure

```{r}
# Final Plotting
s <- ggplot(mydata_ordered, aes(x = group, y = cleaned_value, color = group)) +
  stat_summary(fun = "mean", geom = "bar", fill = "white", linewidth = 1, width = 0.7) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, color = "black") + 
  geom_jitter(width = 0.2, alpha = 0.5, size = 0.8) + 
  
  stat_compare_means(
    comparisons = my_comparisons,
    label = "p.signif",        # This tells R to use stars (*, **) instead of numbers
    method = "wilcox.test",   
    hide.ns = TRUE            # Don't show 'ns' if the difference is not significant
  ) +
  facet_wrap(~measurement, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(
    "Non-COVID-19" = "#4A69BD", 
    "Non-severe"   = "#F6B93B", 
    "Severe"       = "#B71C1C"
  )) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"), # Makes labels clearer
    legend.position = "top",
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank()
  ) +
  labs(title = "Supplementary Figure 1 (Ordered & Cleaned)", y = "Value", x = NULL)

ggsave("SuppFig1_2_2.png", s, width = 14, height = 12, dpi = 300)

s

#Count removed outliers
outlier_summary <- mydata_ordered %>%
  group_by(measurement, group) %>%
  summarise(
    total_samples = n(),
    outliers_removed = sum(is.na(cleaned_value) & !is.na(value)),
    percent_removed = (outliers_removed / total_samples) * 100
  ) %>%
  filter(outliers_removed > 0) # Only show markers that actually had outliers

print(outlier_summary)
```

## Interpretation

The primary difference between the original Supplementary Figure 1 and the reproduced ordered and cleaned figure is the significant improvement in visual clarity and normalization of the y-axis scales. 
In the original figure, extreme outliers (such as an ALT value of around 810 or an AST value near 800) forced the y-axis to extend to very high limits. This "squashed" the bars representing the means, making it difficult to visually distinguish differences between groups. In the cleaned figure, removing these extreme values, allowed the y-axis to make a "zoom in", making the biological trends and differences between the non-COVID19, non-severe and severe groups much more apparent. 
In the original figure, the dots looked more concentrated because the scale was so large. By removing the extreme points, the plot expands the middle range, and that's the reason why the dots now look more spread out. 

Regarding the outlier removal process, was performed per group and per marker using the 1.5*IQR method. This ensured that higher baseline levels in the severe group were not incorrectly categorized as outliers when compared to other groups. Furthermore, a total of 30 distinct group/marker combinations had outliers removed. 

Moreover, the outlier removal process directly influenced the statistical sensitivity of the comparisons between groups:

- Increased Sensitivity: In the cleaned plot, markers like WBC count and CRP show more robust significance levels (appearing as *** or **). Removing extreme values reduced the within-group variance, allowing the statistical tests to more effectively detect the underlying biological differences between severe and non-severe cases.

- Shift in p-values: Some comparisons that were borderline significant or dominated by noise in the original figure now show clearer p-value shifts. For example, the glucose concentration consistently shows a higher concentration in the severe group across both plots, but the cleaned version provides a more precise statistical context for this elevation.

- Persistent non-significance: Despite the "zoom" effect, markers such as monocytes, platelets, and creatinine remain not significant. This confirms that the lack of difference in these markers is a biological reality rather than a result of outlier interference.

The fact that the original paper kept these extreme values (original figure), suggest that they prioritized showing the full clinical range of COVID19, including extreme organ failure, rather than achieving a "normal" statistical distribution. This reproduction demonstrates that standard IQR cleaning focuses on the typical case but can make the remaining variance appear more pronounced when the axis is allowed to scale freely. 

The original paper likely retained extreme values to showcase the full clinical spectrum of COVID-19, including instances of extreme organ failure. However, this reproduction demonstrates that while standard IQR cleaning focuses on the typical patient case, it also makes the remaining variance appear more pronounced when the axis scales freely. Ultimately, cleaning the data did not just remove "dots", it fundamentally changed the resolution of the clinical phenotypes.

# Exercise 3 – Heatmap analysis

This are some considerations or comments regrding the following code that helped me understand better the exercise:

- As biomarkers have very different units, units must be scaled so all of them are compared on a relative basis. This is done in the pheatmap function to do the heatmap
- By grouping by group, allows to see if "severe" patients have a distinct color pattern compared to the "healthy" controls
- Grouping by gender, helps identify if biomarker patterns are influenced by biological sex. 


## Data standardization

```{r}

#install.packages("pheatmap")
library(pheatmap)
library(tidyverse)
library(ggplot2)

# Take cleaned "mydata_ordered" and pivot it back to wide format
mydata_long <- mydata %>%
  select(patient_id_a, group = group_d, sex = sex_g, all_of(ordered_vars)) %>%
  mutate(across(all_of(ordered_vars), ~as.numeric(as.character(.)))) %>%
  pivot_longer(cols = all_of(ordered_vars),
               names_to = "measurement",
               values_to = "value") %>%
  drop_na(value)

heatmap_ready <- mydata_long %>%
  drop_na(value) %>%
  pivot_wider(names_from = measurement, values_from = value) %>%
  drop_na() # Heatmaps cannot handle NA values easily; pheatmap will fail

heatmap_ready <- heatmap_ready %>%
  mutate(
    group = case_when(
      group %in% c(0, 1) ~ "Non-COVID-19",
      group == 2 ~ "Non-severe",
      group == 3 ~ "Severe",
      TRUE ~ NA_character_
    ),
    group = factor(group, levels = c("Non-COVID-19", "Non-severe", "Severe")),
    
    sex = case_when(
      sex == 0 ~ "Female",
      sex == 1 ~ "Male",
      TRUE ~ as.character(sex)
    )
  )

# Prepare Annotation
annotation_col <- heatmap_ready %>%
  select(group, sex) %>%
  as.data.frame()

# Create the Matrix
matrix_to_plot <- heatmap_ready %>%
  select(all_of(ordered_vars)) %>%
  as.matrix()

# Plot
## Convert everything to standard data frames immediately
matrix_df <- as.data.frame(matrix_to_plot)
# Make sure group and sex are in my annotation data
# annotation_col <- heatmap_ready[rows_to_keep, ] %>% 
#   select(group, sex) %>% 
#   as.data.frame()

## Handle the NAs (Outliers)
# Since I have 51 patients in my matrix, we need to make sure 
# the annotation table also has exactly those 51 patients.
# rows_to_keep <- complete.cases(matrix_df)

# matrix_to_plot_clean <- matrix_df[rows_to_keep, ]
# annotation_col_clean <- annotation_col[rows_to_keep, , drop = FALSE]

## Assign matching dummy names
patient_names <- paste0("Patient_", 1:nrow(matrix_to_plot))
rownames(matrix_to_plot) <- patient_names
rownames(annotation_col) <- patient_names

## Check the counts
print(paste("Matrix rows (patients):", nrow(matrix_to_plot)))
print(paste("Annotation rows:", nrow(annotation_col)))
```

## Heatmap construction

```{r}
## Plot
h <- pheatmap(
  t(matrix_to_plot), # Biomarkers in rows
  scale = "row",     # Z-score scaling
  annotation_col = annotation_col,
  show_colnames = FALSE,
  main = "Heatmap: Clinical Biomarkers by Group and Sex"
)

h

ggsave("Heatmap.png", h, width = 14, height = 12, dpi = 300)
```

## Interpretation

The heatmap presents standardized clinical biomarker values across individuals, grouped by disease status and annotated by biological sex. Each row represents a biomarker, while each column corresponds to a patient. Values were standardized (Z-scores), allowing comparison across variables with different measurement units. Warmer colors (red/orange) indicate relatively higher values, whereas cooler colors (blue) indicate lower values compared to the overall mean. The hierarchical clustering of columns reveals a tendency for patients to group according to disease severity rather than sex. Individuals classified as severe COVID-19 cluster together and display a consistent pattern of elevated inflammatory and metabolic markers such as CRP, AST and ALT. In contrast, non-severe and non-COVID-19 patients show relatively lower or more balanced biomarker profiles.
This suggests that disease severity is associated with a distinct biochemical signature.

# Conclusions

Clinical characterization across the three patient groups reveals that severe COVID-19 is defined by systemic inflammatory and metabolic dysregulation. Glucose and CRP are the most significant differentiators, with the Severe group exhibiting markedly higher median levels, elevated glucose indicates a profound physiological stress response, while CRP serves as a critical marker associated with disease progression. Hierarchical clustering in the biomarker heatmap shows that these inflammatory signatures (specifically the elevated CRP, ALT, and AST cluster) emerge independently of sex, as both male and female patients are distributed throughout the severe and non-severe clusters. Immunologically, a decreased lymphocyte count is distinguished between  the severe cohort from non-severe and non-COVID-19 patients, whereas WBC, monocyte and platelet counts remain relatively stable across groups. While liver stress markers (ALT and AST) show slight upward trends in severe cases, suggesting potential multi-organ involvement. Other parameters such as TBIL and DBIL do not show significant variation, indicating that the primary clinical signature of severity in this cohort is driven by acute inflammation and metabolic shifts rather than broad hepatic or biliary failure and this progression appears to follow a consistent pattern across both biological sexes in this dataset. 



# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
